---
title: "ICA 5"
author: "Derien Weatherspoon"
date: "2023-02-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libaries}
attach(mtcars)
library(tidyverse)
```


## Question 1
Examine the qsec field in the dataset. Calculate its mean, standard deviation, and standard error.
The following questions assume that the qsec field is normally distributed. 

```{r Q1}
mean(mtcars$qsec)
sd(mtcars$qsec)

std.error <- function(x){ sd(x)/sqrt(length(x))
}
std.error(mtcars$qsec)
```

## Question 2
(a) Construct a 95% 2-sided confidence interval for the mean of the qsec field.
(b) Then, construct a 99% 1-sided confidence interval (lower limit).
(c) What would the result be for the 99% confidence interval if you used a normal distribution in place of
the t-distribution?

```{r Q2, warning=FALSE}
#a
attach(mtcars)
qsec.model <- lm(qsec ~ 1, mtcars) # mean of qsec
t.test(qsec)$conf.int # 95% confidence interval
#b
confint(qsec.model,level = 0.99) # 99% confidence interval?
#c
sd_q <- sd(mtcars$qsec)
sqrt_q <- sqrt(length(mtcars$qsec))
mean_q <- mean(mtcars$qsec)
new_CI <- qnorm(.99)*sd_q/sqrt_q
lower = mean_q - new_CI
lower
```

## Question 3
How large would the sample size have to be if a 99% 2-sided confidence interval were to be 0.2
seconds wide (given the same standard deviation)?

```{r Q3}

```

## Question 4
Suppose you want to show that the mean value is different from 17 seconds, at a 95% confidence
level. Conduct a Monte Carlo simulation to determine whether you can falsify the null hypothesis.
What is the p-value?


## References

https://www.statology.org/standard-error-of-mean-r/
https://bookdown.org/logan_kelly/r_practice/p09.html

