---
title: "ICA 12"
author: "Derien Weatherspoon"
date: "2023-04-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tseries)
library(forecast)
```


## Question 1

Import the data, convert it to a time series, and then plot it.

```{r Q1}
beer <- read.csv("beer.csv")
beer_ts <- ts(beer[,2], start = c(1956,1), frequency = 12)
plot(beer_ts, main = "Monthly Beer Production", ylab = "Million Barrels")
```

## Question 2

Do additive and multiplicative decompositions of the data. Which one looks better? From the seasonal component, what part of the year typically has the largest beer consumption?

```{r Q2}
beer_additive <- decompose(beer_ts, type = "additive")
beer_multiplicative <- decompose(beer_ts, type = "multiplicative")
plot(beer_additive$seasonal)
plot(beer_multiplicative$seasonal)
```

# Question 3

Create the following forecast models for the data
a. Naïve
b. Seasonal naïve
c. Holt-Winters
d. ARIMA

```{r Q3}
# naive model
beer_naive <- naive(beer_ts)
plot(beer_naive)

# seasonal naive model
beer_snaive <- snaive(beer_ts, h = 12)
plot(beer_snaive)

# holt-winters
beer_hw <- hw(beer_ts, h = 12)
plot(beer_hw)

# automatic arima
beer_arima <- auto.arima(beer_ts)
#plot(beer_arima)
```
# Question 4

Print out the summary of each model. Compare the MAPE’s and MAE’s.

```{r Q4}
summary(beer_naive)
summary(beer_snaive)
summary(beer_hw)
summary(beer_arima)
```

