---
title: "HW 1 Solutions"
author: "Paul Speaker"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(sqldf)
library(tidyverse)

order_details <- read.csv("order_details.csv")
orders <- read.csv("orders.csv")
territories <- read.csv("territories.csv")
regions <- read.csv("regions.csv")
employee_territories <- read.csv("employee_territories.csv")
employees <- read.csv("employees.csv")
customers <- read.csv("customers.csv")
shippers <- read.csv("shippers.csv")
suppliers <- read.csv("suppliers.csv")
products <- read.csv("products.csv")
categories <- read.csv("categories.csv")

```

#1 Perform a sort of orders by employeeID, then by shipVia, and then by freight, 
# for those orders by shipped to France.
Order_France <- sqldf("SELECT * FROM orders WHERE shipCountry = 'France' ORDER BY shipVia, freight")
```{r}
#1 Perform a sort of orders by employeeID, then by shipVia, and then by freight, 
# for those orders by shipped to France.
Order_France <- sqldf("SELECT * FROM orders WHERE shipCountry = 'France' ORDER BY shipVia, freight")
glimpse(Order_France)
```

```{r}
#2 Which shipVia has the largest average cost?
order_cost <-sqldf("SELECT ShipVia, AVG(freight) FROM orders GROUP BY ShipVia")
glimpse(order_cost)


# shipVia 2 has the largest cost.  This corresponds to shipper United Package 
# (since the field had a different name I will accept either).
```

```{r}
#3 Which product category has the highest average UnitPrice?  The Lowest?
average_cost <-sqldf("SELECT CategoryID, AVG(UnitPrice) FROM products GROUP BY CategoryID")
category_costs <- sqldf("Select average_cost.*, categories.categoryName FROM average_cost INNER JOIN categories WHERE average_cost.categoryID = categories.categoryID")
category_costs
# Highest category cost is Meat/Poultry, Lowest is Grain/Cereals


```

```{r}
#4 Which products are supplied by a company in the United States?
USproducts <- sqldf("SELECT products.ProductName FROM products INNER JOIN suppliers WHERE products.SupplierID = suppliers.SupplierID AND suppliers.Country='USA'")
USproducts
```

```{r}
#5 Which shipper is shipping the largest number of units of product?  
# Answer in terms of units; you do not need to consider quantityPerUnit here.
full_orders <- sqldf("SELECT orders.*, order_details.productID, order_details.unitPrice, order_details.quantity, order_details.discount from orders INNER JOIN order_details WHERE orders.orderID = order_details.orderID")
total_shipQ <- sqldf("SELECT SUM(full_orders.quantity), shippers.companyName FROM full_orders INNER JOIN shippers WHERE full_orders.ShipVia = shippers.shipperID GROUP BY full_orders.shipVia")
total_shipQ
# United Package, which is shipping 19,945 units

```

```{r}
#6 Which employee is tied to the most sales revenue?  
# Give the name, not the code, along with the total revenue for the employee.
order_revenue <- sqldf("SELECT *, unitPrice*quantity*(1-discount) as revenue from order_details")
employee_revenue <- sqldf("select orders.employeeID, SUM(order_revenue.revenue) FROM orders INNER JOIN order_revenue WHERE orders.orderID = order_revenue.orderID GROUP BY orders.employeeID")
revenue_by_name <- sqldf("SELECT employee_revenue.*, employees.lastName, employees.firstName FROM employee_revenue INNER JOIN employees WHERE employee_revenue.employeeID = employees.employeeID")
revenue_by_name
# Margeret Peacock has the most revenue

```


```{r}
#7 Find the total revenue for each product category.
product_category <- sqldf("SELECT categories.categoryName, products.ProductID FROM categories INNER JOIN products WHERE categories.CategoryID =  products.CategoryID")
Cat_revenue <- sqldf("SELECT product_category.categoryName, SUM(order_revenue.revenue) FROM product_category INNER JOIN order_revenue WHERE product_category.productID = order_revenue.productID  GROUP BY product_category.categoryName")
Cat_revenue
```


```{r}
#8 Consider the amount of revenue for each customer.  If there were no discounts applied, 
# which customer would see the largest increase in cost?
order_disc <- sqldf("SELECT unitPrice*quantity*discount as revenue_disc, orders.customerID FROM order_details INNER JOIN orders WHERE order_details.orderID = orders.orderID")
total_disc <- sqldf("SELECT customers.companyName, SUM(order_disc.revenue_disc) FROM customers INNER JOIN order_disc WHERE customers.customerID = order_disc.customerID GROUP BY customers.companyName ORDER BY SUM(order_disc.revenue_disc)")
# Answer is Save-a-lot Markets
```

```{r}
#9 Which order(s) has the most number of items (and how many)?  Give the orderID for this one.
item_count = sqldf("SELECT orderID, COUNT(productID) FROM order_details GROUP BY orderID ORDER BY count(productID) DESC")
glimpse(item_count)
# OrderID 11077 has 25 items

```
# 10 Create a new field called "InventoryOrderRatio" which is, for each product, the UnitsinStock (the inventory) for the product (across all customers) divided by the quantity ordered for that product.  
# A high value represents sufficient product in stock, while a low number represents products that are in 
# danger of running out.  What 3 products are most in danger of running out?

```{r}
inventory_data <- sqldf("SELECT products.ProductName, products.UnitsInStock, SUM(order_details.quantity) AS total_q 
                        FROM products 
                        INNER JOIN order_details 
                        ON products.productID = order_details.productID 
                        GROUP BY order_details.productID")

inventory_data$quant <- as.numeric(inventory_data$total_q)

inventory_ratio <- sqldf("SELECT ProductName, UnitsInStock/quant AS InventoryOrderRatio 
                          FROM inventory_data 
                          WHERE UnitsInStock > 0 AND total_q > 0 
                          ORDER BY UnitsInStock/quant")

inventory_ratio
# 3 smallest inventory_ratios are for Sir Rodney's Scones, Scottish Longbreads, and Rogede sild
# ok if ones where inventory are 0 are included--these will be a different list.
```

```{r}
# 11 A recommender engine looks at which pairs of products tend to be bought by the same # customer, so that if a customer buys one, the recommender engine will recommend they buy the other.  Find which product pairs are most likely to be bought by the same customer. 
# (Hint: this will require a creative join with the orders data)
full_orders <- sqldf("SELECT orders.customerID, order_details.productID
                     FROM orders
                     INNER JOIN order_details
                     ON orders.orderID = order_details.orderID")

fo2 <- sqldf("SELECT customerID, productID AS prod1 FROM full_orders")
order_combos <- sqldf("SELECT fo2.customerID, fo2.prod1, full_orders.productID AS prod2
                      FROM fo2 
                      INNER JOIN full_orders 
                      ON fo2.customerID = full_orders.customerID 
                      WHERE fo2.prod1 > full_orders.productID")

combo_count <- sqldf("SELECT prod1, prod2, COUNT(customerID) FROM order_combos 
                     GROUP BY prod1, prod2 
                     ORDER BY COUNT(customerID) DESC")
glimpse(combo_count)
# Most common order pair is product ID 31 and 56, corresponding to Gorgonzola Telino and Gnocchi di nonna Alice
# Gnocchi with a Gorgonzola sauce is amazing.

```

