---
title: "Diamonds are forever"
author: "Hannah Striebel - group 3"
date: "M3 ICA1"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    number_sections: yes
    toc_float: yes
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)
```

# The data


Package `tidyverse` is a set of packages that work in harmony. We will use
the `ggplot2` package to produce our data visualizations. This package is part
of the `tidyverse` package. As we move forward, we will utilize some of the
other packages loaded via `tidyverse`.


```{r packages}
library(tidyverse)
```

The ggplot2 package comes with a data set called `diamonds`. 
Let's look at it below. 
To obtain further details type `?diamonds` in your console window.

```{r}
glimpse(diamonds)
```

**Variable codebook**

The dataset containing the prices and attributes of nearly 54,000 diamonds includes
the following variables:

    *price*
    price in US dollars (\$326--\$18,823)
    
    *carat*
    weight of the diamond (0.2--5.01)
    
    *cut*
    quality of the cut (Fair, Good, Very Good, Premium, Ideal)
    
    *color*
    diamond colour, from D (best) to J (worst)
    
    *clarity*
    a measurement of how clear the diamond is (I1 (worst), SI2, SI1, VS2, VS1, VVS2, VVS1, IF (best))
    
    *x*
    length in mm (0--10.74)
    
    *y*
    width in mm (0--58.9)
    
    *z*
    depth in mm (0--31.8)
    
    *depth*
    total depth percentage = z / mean(x, y) = 2 * z / (x + y) (43--79)
    
    *table*
    width of top of diamond relative to widest point (43--95)
    
# The investigation

**What influences the price of a diamond?**

As a quick primer on diamond pricing, watch the following video on 'the 4C's of Diamonds.'
https://www.youtube.com/watch?v=dFiG3ckNCIY&feature=emb_logo

# Exploring the data

1. Which of the 4C's do you predict most influences the price of a diamond? The least? 
Provide a 1-2 sentence explanation of your opinion.

I predict that cut is going to influence the price of the dimaond most, as this
has a great influence on how the diamond refracts light and thus how "sparkly"
it is. I predict color will have the least influence on price, as color seems
more like a subjective preference than an objective measure of quality.

2. The data set `diamonds` is stored in R as a tibble. This allows for a 
convenient way to view the data frame in the console. Type `diamonds` in your
console to see.

Let's briefly explore the distribution of some of the variables stored in `diamonds`.

3. a. Variables cut, color, and clarity are all factors. Use the function `levels` to see the levels of each variable. They are sorted from worst to best. Use the `table` command to determine how many cases fall into each level of cut Do the same for color and clarity.
```{r}
levels(diamonds$cut)
levels(diamonds$color)
levels(diamonds$clarity)

table(diamonds$cut)
table(diamonds$color)
table(diamonds$clarity)
```


   b. Because `price` is quantitative, the `table` command won't provide a useful summary of the distribution of observed cases for this variable. Use the `summary` function to get a sense of this variable's distribution.
   
```{r}
summary(diamonds$price)
```
   
   
   c. Add a new variable to `diamonds` called `price.per.carat` that represents the price per carat. Are all diamonds priced the same per carat? If not, how much do these rates vary?
   
```{r}
diamonds$price.per.carat <- diamonds$price / diamonds$carat
summary(diamonds$price.per.carat)
```
No, they are not all priced the same per carat - they range from about one
thousand to almost 18 thousand dollars per carat. The mean is slightly greater
than the median, which means there are more outliers that are exorbitantly
expensive than underpriced.
  
Write 1-2 sentences hypothesizing what might cause a diamond to fetch a higher price per carat than others.

If one diamond's cut, color, or clarity of are higher quality, then that diamond
would be more expensive, even if the two diamonds are the same carat.


In contrast to (3a) and (3b), which ask about the distribution of a single variable, exercise (3c) is the first to begin looking at the *relationship* between two variables - in this case, a diamond's `price` and `carat`.

Let's investigate relationships further. **What is the relationship between a diamond's price and its cut/color?**

4. a. Remember that a diamond's cut can be too shallow
or too steep, and either will cause it to sparkle less dramatically under bright light.

Summarize and compare the prices of Fair (the worst cut) and Ideal (the best cut) diamonds.

```{r diamond cutcomparison1}
summary(diamonds$price[diamonds$cut == "Fair"])
summary(diamonds$price[diamonds$cut == "Ideal"])

```

How unexpected! Many of the quantile prices for diamonds rated
Fair exceed those rated Ideal (every value of the five-number summary apart from the max price]. 
This is probably the exact opposite of what you may have expected.

b. Summarize and compare the prices of D (the best color) diamonds and J (the worst color) diamonds.

```{r diamond colorcomparison}
#use the summary function on price for the particular colors.
summary(diamonds$price[diamonds$color == "D"])
summary(diamonds$price[diamonds$color == "J"])

```

Another unexpected result! The summary statistics show that many poorly colored diamonds fetch higher prices than ideally colored diamonds at the same percentile ranking (i.e., the 75th most expensive perfectly-colored diamond is much cheaper than the 75th most expensive poorly-colored diamond). How can this be?

5. **What influences a diamond's price? ** In 1-2 sentences, hypothesize as to why the distribution of poorly cut diamonds trends more expensive than those that are perfectly cut.

I predict that clarity must be the most influential on price. Thus, poorly
colored or cut diamonds with high clarity are priced much higher than well
cut and colored diamonds with low clarity.

# Visualizations with ggplot

Visualizations are an excellent tool for exploring the distributions of variables - their patterns, their variability,
and how those attributes are related to those of other variables. The exercises below ask you to first recreate a number
of visualizations we believe will help to answer our initial question **What influences a diamond's price?** along with
several others along the way.

## Exercises {.tabset .tabset-fade .tabset-pills}

5. The code to create each plot below is given. Try to understand what the associated code is doing. As we start delving into Data Visualization you will gain experience and can recreate these plos on your own.

6. Consider Plot 1. Notice how the distribution of `carat` vs. `price` has clustered vertical lines at 1, 1.5, and 2 carats. It seems suspicious that naturally occurring diamonds would appear at these values more often than others (such as 0.9, 1.4, or 1.95). Provide a plausible explanation for why our data on ~54,000 diamonds might display this pattern.

As humans, we really like to work with whole numbers. So if a diamond is close
to a whole number carat, it makes sense that the jeweler would round to the
nearest whole number. In addition, it looks like there are very few between
about 1.75 carat and 2 carat, which likely means that jewelers will round up to
the nearest whole carat if it is close.

7. Consider Plot 2, which displays the relationship between `carat`, `color`, and `price`, and remember that color is coded across letters D to J, from most desirable to least desirable. We earlier discovered that poorly colored diamonds tend to fetch higher prices than ideally colored diamonds. Write 1-2 sentences describing how this plot helps to explain this counterintuitive finding.

It appears that higher carat diamonds are more often lower quality in color.
Thus, if carat has a higher impact on price than color, then the lower quality
colored diamonds will be more expensive than the higher quality colors.

8. You'll notice Plots 3 and 4 display the relationship between the same four variables, `carat`, `color`, `price`, and `cut`. Which of the two plots is more useful when analyzing which factors play a role in the overall price of a diamond? Justify your opinion in 2-3 sentences.

Plot 4 is much easier to see the relationships because it plots them in five
different graphs. Plot 3 is very difficult to read because there is too much
information all on top of each other. I can't tell which icons for cut are where
on the graph becuase the dots for color cover them up. However, from Plot 4, we
can compare more easily across different colors, cuts, and carats becuase they
are more clearly displayed.

9. Finally, consider Plot 5, which considers the final of the '4 C's' of diamond pricing, clarity. You should notice that the clearest of diamonds (rating IF) include many more ideal cuts than the foggiest of diamonds (rating l1). Hypothesize as to why this might be? 

It probably isn't worth the effort to cut a cloudy diamond perfectly - the fact
that it is cloudy means it won't be as sparkly even if you cut it perfectly, so
it doesn't make sense to waste time and resources if cutting it perfectly won't
improve the price.

<font color="black">

### Plot 1 {-}

</font>

```{r plot1, echo=FALSE, cache=TRUE}
diamond.plot <- ggplot(data = diamonds, aes(x = carat, y = price))
diamond.plot + geom_point()
```

### Plot 2 {-}

```{r plot2, echo=FALSE, cache=TRUE}
diamond.plot <- ggplot(data = diamonds, aes(x = carat, y = price, color = color))

diamond.plot + geom_point() + 
  labs(y = "Price ($)", 
       x = "Carat", title = "Diamond Carat vs Price", 
       subtitle = "data from ggplot2")
```

### Plot 3 {-}

```{r plot3, echo=FALSE, cache=TRUE}
diamond.plot <- ggplot(data=diamonds, 
                       aes(x = carat, y = price,  shape = cut, colour = color))
diamond.plot + geom_point()+ 
  labs(y = "Price ($)", x = "Carat", 
       title = "Diamond Carat vs Price", subtitle = "data from ggplot2")
```

### Plot 4 {-}

```{r plot4, echo=FALSE, cache=TRUE}
diamond.plot <- ggplot(data = diamonds, aes(x = carat, y = price, colour = color))

diamond.plot + geom_point() + facet_wrap(~cut) + 
  labs(y = "Price ($)", x = "Carat")
```

### Plot 5 {-}

```{r plot5, echo=FALSE, cache=TRUE}
diamond.plot <- ggplot(data = diamonds, aes(x = carat, y = price, colour = cut))

diamond.plot + geom_point() + facet_wrap(~clarity, nrow = 2) + 
  labs(y = "Price ($)", x = "Carat")
```

# Wrap up

In this in-class activity, we used visualization to investigate the relationships between two or more variables. You likely noticed that apparent relationships between two variables (for instance, that poorer colored diamonds tended to be more expensive) were sometimes influenced by the interaction with a third, unobserved variable. Such unobserved variables are often considered *confounding variables*. 

In this case, a diamond's `carat` (or size) is very strongly associated with its `price`. Additionally, the larger a diamond grows, the more unlikely it is to maintain ideal color or clarity, and the more difficult it is for a jeweler to cut it perfectly. This helps to explain our initial finding that poorly colored or cut diamonds were more expensive - often, they were just bigger. This is one example of the power of data visualization, which helps make intelligible patterns in data that might not otherwise make sense.

# References

1. Grolemund, G., & Wickham, H. (2019). R for Data Science. 
   https://r4ds.had.co.nz/
