---
title: "ICA 7"
author: "Derien Weatherspoon"
date: "2023-03-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1
Define these matrices in R (using whichever technique you prefer)

```{r}
# define matrices
A <- matrix(c(5,-1,5,-1,10,1,5,1,-3), nrow = 3, ncol = 3)
B <- matrix(c(5,1,3,1,-2,0), nrow = 2, ncol = 3)
C <- matrix(c(2,1,-3,4), nrow = 2, ncol = 2)
x <- matrix(c(5,1,4), nrow = 3, ncol = 1)

```

## Question 2
Compute CB


```{r}
CB <- C %*% B
CB
```

## Question 3
Compute Ax

```{r}
Ax <- A %*% x
Ax
```

## Question 4
Find det(C)

```{r}
det(C)
```

## Question 5
Calculate BtB and BBt. What kind of matrices are these results?
 
```{r}
Bt <- t(B)
BtB <- Bt %*% B
BtB

BBt <- B %*% Bt
BBt
```


## Question 6
Compute xtx. Note that this is the square of the norm of x.

```{r}
xt <- t(x)
xtx <- xt %*% x
xtx
```

## Question 7
Solve the following for the unknown y: Ay = x.

```{r}
y <- solve(A,x)
y
```

## Question 8
Find the eigenvalues and eigenvectors of A. Note that the eigenvectors are perpendicular (this
can be checked by seeing that the inner product is zero).

```{r}
eigen(A)$values
eigen(A)$vectors
```

## Question 9
Find the eigenvalues and eigenvectors of A2. How do the eigenvalues compare to that of A?

```{r}
A2 <- A %*% A
eigen(A2)$values
eigen(A2)$vectors
```

## Question 10
Take a vector z = (1, -3, 4). Re-write z using the eigenvectors of A as the new basis.

```{r}
z <- c(1,-3,4)

solve(eigen(A)$vectors, z) * z
```


