---
title: "SS23 STT481: Lab: Subset Selection and Shrinkage Methods"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

``` {r echo=TRUE}
library(ISLR)
data("Hitters")
head(Hitters)
```
**Goal**: We wish to predict a baseball player's `Salary` on the basis of various statistics associated with performance in the previous year.

**Preprocessing**: remove missing observations.
``` {r echo=TRUE}
which(is.na(Hitters$Salary))   # it returns the index of missing Salary's
head(Hitters[which(is.na(Hitters$Salary)),]) # Salary's are NA's
Hitters <- Hitters[!is.na(Hitters$Salary),] # We remove those rows which contain NA's
which(is.na(Hitters$Salary)) 
```

# Subset Selection Methods

## Best Subset Selection
``` {r echo=TRUE}
library(leaps)
regfit.full <- regsubsets(Salary ~ . , data = Hitters)
#
summary(regfit.full)
```
The default of `nvmax` option in `regsubsets` so it only reports up to the best 8-variable model. We can change it to `nvmax=19` to fit up to a 19-variable model. Note when `nvmax` is moderate large (say, 50), it might take a long time to fit the model. 

``` {r echo=TRUE}
regfit.full <- regsubsets(Salary ~ . , data = Hitters, nvmax = 19)
reg.summary <- summary(regfit.full)
names(reg.summary)
```
It returns training errors and estimated test errors.

``` {r echo=TRUE}
reg.summary$rsq 
```
As expected, $R^2$ statistic increases monotonically as more variables are included.

Plotting RSS, adjusted R2, Cp, and BIC for all of the models at once will help us decide which model to select.
``` {r echo=TRUE}
par(mfrow=c(2,2))
plot(reg.summary$rss ,xlab="Number of Variables ", ylab="RSS", type="l")
plot(reg.summary$adjr2 ,xlab="Number of Variables ", ylab="Adjusted RSq", type="l")
which.max(reg.summary$adjr2)
points(11,reg.summary$adjr2[11], col = "red", cex = 2, pch = 20)
plot(reg.summary$cp, xlab = "Number of Variables ", ylab = "Cp", type = "l")
which.min(reg.summary$cp)
points(10,reg.summary$cp[10], col = "red", cex = 2, pch = 20)
which.min(reg.summary$bic)
plot(reg.summary$bic, xlab = "Number of Variables ", ylab = "BIC", type = "l")
points(6,reg.summary$bic[6], col = "red", cex = 2, pch = 20)
```

The `regsubsets()` function has a built-in `plot()` command which can be used to display the selected variables for the best model with a given number of predictors, ranked according to the BIC, Cp, adjusted R2, or AIC.

```{r echo=TRUE}
plot(regfit.full,scale="r2")
plot(regfit.full,scale="adjr2")
plot(regfit.full,scale="Cp")
plot(regfit.full,scale="bic")
```

We can use the `coef()` function to see the coefficient estimates associated with this model.
```{r echo=TRUE}
coef(regfit.full, 6)
```

## Forward and Backward Stepwise Selection
``` {r echo=TRUE}
regfit.fwd <- regsubsets(Salary ~., data = Hitters, nvmax = 19, method = "forward")
#summary(regfit.fwd)
regfit.bwd <- regsubsets(Salary ~., data = Hitters, nvmax = 19, method = "backward")
#summary(regfit.bwd)
```

Similarly, we can plot RSS, adjusted R2, Cp, and BIC for all of the models seleced by forward stepwise selection.
``` {r echo=TRUE}
par(mfrow=c(2,2))
reg.fwd.summary <- summary(regfit.fwd)
plot(reg.fwd.summary$rss ,xlab="Number of Variables ", ylab="RSS", type="l")
plot(reg.fwd.summary$adjr2 ,xlab="Number of Variables ", ylab="Adjusted RSq", type="l")
which.max(reg.fwd.summary$adjr2)
points(11, reg.fwd.summary$adjr2[11], col = "red", cex = 2, pch = 20)
plot(reg.fwd.summary$cp, xlab = "Number of Variables ", ylab = "Cp", type = "l")
which.min(reg.fwd.summary$cp)
points(10,reg.fwd.summary$cp[10], col = "red", cex = 2, pch = 20)
which.min(reg.fwd.summary$bic)
plot(reg.fwd.summary$bic, xlab = "Number of Variables ", ylab = "BIC", type = "l")
points(6,reg.fwd.summary$bic[6], col = "red", cex = 2, pch = 20)
```

The selected models identified by forward, backward stepwise selection and best subset selection are different.
``` {r echo=TRUE}
coef(regfit.full, 7)
coef(regfit.fwd, 7)
coef(regfit.bwd, 7)
```

## Choosing Among Models using Cross-Validation
We just saw that it is possible to choose among a set of models of different sizes using Cp, BIC, and adjusted R2. We will now consider how to do this using the cross-validation approach.

Because there is no `predict()` method for `regsubsets()`, we can write our own predict method.
``` {r echo=TRUE}
predict.regsubsets <- function (object, newdata , id, ...){
  form <- as.formula(object$call[[2]])  # formula of null model
  mat <- model.matrix(form, newdata)    # building an "X" matrix from newdata
  coefi <- coef(object, id = id)        # coefficient estimates associated with the object model containing id non-zero variables
  xvars <- names(coefi)            # names of the non-zero coefficient estimates
  return(mat[,xvars] %*% coefi)    # X[,non-zero variables] %*% Coefficients[non-zero variables]
}
```
For example, suppose we want to predict the newdata below using the models containing 6 variables selected by best subset selection.
``` {r echo=TRUE}
newdata <- Hitters[1:2, ] ## I made up a newdata
newdata[,"Hits"] <- newdata[,"Hits"] + rep(60,2) ## I made up a newdata
newdata[,"HmRun"] <- newdata[,"HmRun"] + rep(30,2) ## I made up a newdata
predict.regsubsets(regfit.full, newdata = newdata, id = 6)
```

Now we are ready to perform a 10-fold cross-validation.
``` {r echo=TRUE}
fold.index <- cut(sample(1:nrow(Hitters)), breaks=10, labels=FALSE)
cv.error.best.fit <- rep(0,19)
for(i in 1:19){   # try different numbers of variables
  cat("i=", i,"\n")
  error <- rep(0, 10)
  for (k in 1:10){
    Hitters.train <- Hitters[fold.index != k,]
    Hitters.test <- Hitters[fold.index == k,]
    true.y <- Hitters.test[,"Salary"]
    best.fit <- regsubsets(Salary ~ ., data = Hitters.train, nvmax = 19)
    pred <- predict(best.fit, Hitters.test, id = i)
    error[k] <- mean((pred - true.y)^2)
  }
  print(mean(error))
  cv.error.best.fit[i] <- mean(error)
}
```

``` {r echo=TRUE}
par(mfrow=c(1,1))
plot(cv.error.best.fit, type = "b")
points(which.min(cv.error.best.fit), cv.error.best.fit[which.min(cv.error.best.fit)], 
       col = "red", cex = 2, pch = 20)
```

We see that cross-validation selects an `r which.min(cv.error.best.fit)`-variable model. We now perform best subset selection on the full data set in order to obtain the `r which.min(cv.error.best.fit)`-variable model.

``` {r echo=TRUE}
reg.best <- regsubsets(Salary ~ ., data = Hitters, nvmax = 19)
coef(reg.best, which.min(cv.error.best.fit))
```
**Notice that the data = Hitters**, which is the full data set. Don't miss this step after performing cross-validation methods.

# Shrinkage Methods
We will use the `glmnet` package in order to perform ridge regression and the lasso.  The main function in this package is `glmnet()`, which can be used to fit ridge regression models, lasso models, and more. This function has slightly different syntax from other model-fitting functions that we have encountered thus far. In particular, we must pass in an `x` matrix as well as a `y` vector, and we do not use the `y ~ x` syntax. In order to do so, we use `model.matrix()` to create `x`, which not only produces a matrix corresponding to the 19 predictors but it also automatically transforms any qualitative variables into dummy variables.

``` {r echo=TRUE}
X <- model.matrix(Salary ~., Hitters)[,-1]  # the first column is the intercept
y <- Hitters$Salary
```

## Ridge Regression
``` {r echo=TRUE}
library(glmnet)
ridge.mod <- glmnet(X, y, alpha = 0)  # 0 indicates ridge regression
```
Of course you can also assign the values of $\lambda$ to perform the ridge regression.
``` {r echo=TRUE}
grid <- 10^seq(10, -2, length = 100)
ridge.mod <- glmnet(X, y, alpha = 0, lambda = grid)
```

Associated with each value of $\lambda$ is a vector of ridge regression coefficients, stored in a matrix that can be accessed by `coef()`. In this case, it is a $20\times 100$ matrix, with 20 rows (one for each predictor, plus an intercept) and 100 columns (one for each value of $\lambda$).

``` {r echo=TRUE}
dim(coef(ridge.mod))
```


``` {r echo=TRUE}
ridge.mod$lambda[50]
coef(ridge.mod)[,50] # this indicates the coefficent estimates when lambda = 11498
```

In general, we would like to use cross-validation to choose the tuning parameter $\lambda$. We can do this using the built-in cross-validation function, `cv.glmnet()`. By default, the function performs 10-fold cross-validation, though this can be changed using the argument `nfolds`. 
``` {r echo=TRUE}
cv.out <- cv.glmnet(X, y, alpha = 0, nfolds = 10)
plot(cv.out)
bestlam <- cv.out$lambda.min
bestlam
```

We can examine the coefficient estimates when $\lambda=$`bestlam` using `coef()`.
``` {r echo=TRUE}
coef(ridge.mod, s = bestlam)
```
As expected, none of the coefficients are zero—ridge regression does not perform variable selection!

Again, I made the following new data and suppose I want to predict their responses (Salary).
``` {r echo=TRUE}
newdata <- Hitters[1:2, ] ## I made up a newdata
newdata[,"Hits"] <- newdata[,"Hits"] + rep(60,2) ## I made up a newdata
newdata[,"HmRun"] <- newdata[,"HmRun"] + rep(30,2) ## I made up a newdata
test.X <- model.matrix(Salary ~., newdata)[,-1] # same form as X
ridge.pred <- predict(ridge.mod, s = bestlam, newx = test.X)
ridge.pred
```

## Lasso
In order to fit a lasso model, we once again use the `glmnet()` function; however, this time we use the argument `alpha=1`.
``` {r echo=TRUE}
library(glmnet)
lasso.mod <- glmnet(X, y, alpha = 1, lambda = grid)
```

Again, we use cross-validation to choose the tuning parameter $\lambda$.
``` {r echo=TRUE}
cv.out <- cv.glmnet(X, y, alpha = 1)
plot(cv.out)
bestlam <- cv.out$lambda.min
coef(lasso.mod, s = bestlam)
```
Many of the coefficient estimates with the best $\lambda$ are exactly zero!

Again, suppose I want to make prediction using the best $\lambda$.
``` {r echo=TRUE}
lasso.pred <- predict(lasso.mod, s = bestlam, newx = test.X)
lasso.pred
```