---
title: "ICA 4"
author: "Derien Weatherspoon"
date: "2023-02-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Question 1:
Take a probability distribution p(x) = 0.5sin(x), 0 < x < pi.

CDF: 0.5(-cos(x)+ 1)

a) Create the quantile function for this distribution. (be careful with finding the cdf)
b) Use the quantile function to create a simulation of this distribution, using n = 100,000. 
Plot the simulation in a histogram. Does the shape match the original distribution?

```{r Q1 a}

#integrand is 0.5sin(x), bounds from 0 to x
x = seq(0,pi,0.05)
n <- 100000
integral <- function(x) {
  result <- 0.5*sin(x)
  return(result)
}

cdf <- function(x){
  result <- 0.5*(-cos(x)+1)
  return(result)
}
q <- function(x){
  result <- acos(-2*x+1)
}

```

```{r Q1 b}
q(runif(n))
hist(q(runif(n)))
```
## Question 2:

In this problem, we will calculate the following familiar integral:
(integral from 0 to pi, sin(x)dx)

a) Calculate the integral by hand

integrand(sin(x)dx) = -cos(x) = [-cos(x)]^pi_0 = 2 <- ANSWER

b) Do 100 iterations of the integral value using standard Monte Carlo, using n = 100,000.
What are the mean and standard deviation of the 100 iterations? (for i in 1:100)

c) Do 100 iterations of the integral value using Acceptance/Rejection Monte Carlo, 
using n = 100,000. What are the mean and standard deviation of the 100 iterations?
d) How do the values compare between (b) and (c)?

```{r Q2 b, echo=FALSE}

for (i in 1:100) {
 ab <- (q(runif(n)))
}
mean(ab)
sd(ab)
```

```{r Q2 c}

```


