---
title: "ICA 11"
author: "Derien Weatherspoon"
date: "2023-04-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r read data and libraries}
housing <- read.csv("housing.csv")
library(GGally)
library(ggplot2)
```

## Question 1: Using "housing.csv", plot the data by longitude and latitude. What data are we looking at?

```{r Q1}
plot(housing$longitude, housing$latitude)
```
The graph is pretty much the state of California.

## Question 2: Trying different numbers of clusters, cluster the data by these location variables. Visually inspect the result (use cluster id as a color). What seems like an optimal number of clusters

```{r Q2}
housing_scaled <- scale(housing[,1:2])
housing_km <- kmeans(housing_scaled, centers = 6)
plot(housing$longitude, housing$latitude, col = housing_km$cluster)
```
Optimal number of clusters seems to be 6.

## Question 2: The rest of the questions are working with the optimal cluster number.
#### a. How big are the different clusters?

The clusters get smaller as you go from top to the bottom. They are not all evenly divided either.

#### b. Filter the data so you just have the largest cluster.

```{r}
housing_cluster <- cbind(housing, housing_km$cluster)
```

#### c. Do a second clustering on this largest cluster, using age, population, and income (this is an example of hierarchical clustering). Is there a natural cluster number for this one?

```{r }

```

#### d. For this largest cluster, create a linear regression model to predict median house value.
#### e. Duplicate the model for the overall dataset. How do the errors compare?
```{r Q3}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
